A general and efficient representation of Ancestral Recombination Graphs

### 2022-05-18 paper.tex
Compressed inference methods section

\subsection*{Parsimony and heuristics}
The problem of reconstructing ARGs for samples of recombining sequences has been of significant
 interest essentially since the ARG was first defined. Early methods focussed on finding the most
 \emph{parsimonious} ARGs, i.e.\ those minimising the number of posited recombination events in
 the history of the sample \citep{hein1990reconstructing}. Two main approaches to explicitly reconstructing
 parsimonious ARGs then emerged: building the ARG either \emph{backwards-in-time} or \emph{along-the-genome}.
 The former approach, introduced by \citet{lyngso2005minimum} and implemented in the program
 \texttt{Beagle}, starts with the input data matrix and reduces it through row and column operations
 corresponding to coalescence, mutation and recombination events, until the empty matrix is reached;
 the corresponding sequence of events can then be used to construct an ARG from the bottom up. The
 latter approach, introduced by \citet{song2003parsimonious} and first implemented in the program
 \texttt{RecMinPath} \citep{song2005constructing}, starts with an initial local tree at some starting
 position, and refines the tree moving left and right along the genome, changing the topology of the
 tree when necessary through \emph{subtree prune and regraft} (SPR) operations (corresponding to
 recombination events).

Reconstructing a most parsimonious ARG for a given data set is NP-hard \citep{wang2001perfect},
and in order to allow for larger input datasets, subsequent parsimony-based methods have resorted
to various relaxations and heuristics. Those utilising variants of the backwards-in-time approach
include the programs \texttt{SHRUB} \citep{song2005efficient}, \texttt{GAMARG} \citep{thao2019hybrid},
and \texttt{KwARG} \citep{ignatieva2021kwarg}, and the method described by \citet{wu2008association}.
Tools constructing genealogies along-the-genome include \texttt{RecPars} \citep{hein1993heuristic},
\texttt{RENT} \citep{wu2011new} and its sequel \texttt{RENT+} \citep{mirzaei2017rent}. Other approaches
include \texttt{TARGet} \citep{camara2016inference}, which implements a method based on topological
data analysis. The scalability of these methods varies substantially based on the heuristics in use
and the underlying data structure, with the most efficient tools limited to analysing hundreds of
sequences.

A number of heuristic methods have also been developed which do not (explicitly) focus on reconstructing
the most parsimonious ARGs, but rather aim for computational efficiency while seeking to achieve good
accuracy of the inferred genealogies. Such methods include \texttt{Margarita} \citep{minichiello2006mapping}
and the approach of \citet{parida2008estimating}; the more recently developed \texttt{Relate}
\citep{speidel2019method} and \texttt{tsinfer} \citep{kelleher2019inferring} use a combination of both
backwards-in-time and along-the-genome approaches, and can be practically used on megabase scale data
and tens to hundreds of thousands of samples under human-like parameters.
Another class of methods has recently emerged that use a \emph{threading} approach, constructing ARGs
by adding sequences to the graph one-by-one. Originally introduced by \citet{rasmussen2014genome} as a
way of generating MCMC proposals for ARGs, a similar idea was leveraged in \texttt{ARGneedle}
\citep{zhang2021biobank} to reconstruct ARGs for biobank-scale data.
% AI: just to note, these papers do not all agree in their definition of an ARG.
% I guess this is one of the points of the present paper, but should we mention this fact?
% Should we also mention that most of these infer just the topology, but some also the times?

Espalier~\citep{rasmussen2022espalier} uses Maximum Agreement Forests to
minimise the differences between local trees estimated using other methods.

\subsection*{Model-based methods}
An alternative approach is to treat the ancestry as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
These methods invariably operated on a representation of the ``little ARG'', typically
generated until the grand MRCA. They are extremely computationally expensive,
and applicable to at most hundreds of samples consisting of tens of kilobases with
human-like parameters.

Due to the computational expense of sampling ``little ARGs'', state-of-the-art
Monte Carlo methods typically rely on cheaper, approximate models.
The most prominent examples are
\texttt{ARGWeaver}~\citep{rasmussen2014genome,hubisz2020mapping,hubisz2020inference}
and \texttt{Arbores} \citep{heine2018bridging}. The former also utilises a time
discretisation approximation, but scales to dozens of mammal-like genomes.
The \texttt{ClonalOrigin} method \citep{didelot2010inference,
medina2020speeding} uses the \texttt{ClonalFrame}
model and data structure, though it is not as scalable as \texttt{ARGWeaver}.

A very recent MCMC method called \texttt{ARGInfer} has resulted in the first
improvement in the scalability of Monte Carlo methods for the exact ARG model
by making use of the tree sequence representation of marginal trees, suitably enriched
to facilitate the computations needed for MCMC sampling
\citep{mahmoudi2022bayesian}.
While not competitive with the scalability of \texttt{ARGWeaver}, \texttt{ARGInfer} extends
the feasible range of sequence lengths by at least an order of magnitude when
compared with methods which sample realisations of the exact ``little ARG''.

>>>>

The problem of reconstructing ARGs for samples of recombining sequences has been of 
interest since the ARG was first defined. Early methods focused on finding \emph{parsimonious} ARGs,
i.e.\ those with a minimal number of recombination events \citep{hein1990reconstructing}.
Two main approaches emerged: \emph{backwards-in-time} \citep{lyngso2005minimum} and
\emph{along-the-genome} \citep{song2003parsimonious, song2005constructing}.
The former starts with a data matrix and reduces it to an empty matrix through row and column operations
corresponding to coalescence, mutation, and recombination events, which construct
an ARG from the bottom up. The latter begins from an initial local tree at a single focal site. Moving the 
focal site along the genome changes the local tree via a \emph{subtree prune and regraft} (SPR) operation
whenever a recombination is traversed. Methods based the backwards-in-time approach are described by
\citet{song2005efficient, wu2008association, thao2019hybrid, ignatieva2021kwarg},
while \citet{hein1993heuristic, wu2011new, mirzaei2017rent} describe along-the-genome methods.
\citet{rasmussen2022espalier} focuses on parsimonious fusion of local trees into an ARG, while
\citet{camara2016inference} is based on topological data analysis.

Reconstructing a most parsimonious ARG for a given data set is NP-hard \citep{wang2001perfect},
so parsimony-based methods resort to heuristics and are limited to analysing hundreds of sequences.
Hence, a number of methods aim to balance computational efficiency with reconstruction of ``reasonable",
rather than parsimonious ARGs
\citep{minichiello2006mapping, parida2008estimating, kelleher2019inferring,  speidel2019method, zhang2021biobank}.
The latter three methods, as well as the parsimony-based method of \citet{rasmussen2022espalier},
can be used on megabase scale data and hundreds of thousands of samples under human-like parameters.

% AI: just to note, these papers do not all agree in their definition of an ARG.
% I guess this is one of the points of the present paper, but should we mention this fact?
% Should we also mention that most of these infer just the topology, but some also the times?

An alternative approach is to treat the ARG as a latent parameter to be averaged out
by Monte Carlo methods, based either on importance sampling
\citep{griffiths1996ancestral, fearnhead2001estimating, jenkins2011inference}
or MCMC \citep{kuhner2000maximum, nielsen2000estimation, wang2008bayesian, fallon2013acg}.
These methods operate on representations of the ``little ARG'', and are extremely computationally
expensive, being applicable to at most hundreds of samples consisting of tens of kilobases with
human-like parameters. State-of-the-art methods rely on cheaper, approximate models
\citep{didelot2010inference, heine2018bridging, hubisz2020mapping,hubisz2020inference, medina2020speeding}. 
The most scalable method, \texttt{ARGWeaver} \citep{rasmussen2014genome}, can be applied to dozens of mammal-like
genomes \citep{hubisz2020inference}.

Recently, \citet{mahmoudi2022bayesian} have improved the scalability of exact MCMC for ARGs by 
leveraging the tree sequence representation of marginal trees, suitably enriched to facilitate the
evaluation of sampling probabilities. While not competitive with \texttt{ARGWeaver}, their method extends
the feasible range of sequence lengths by at least an order of magnitude.
